---
title: "Bike ride share success - case study"
author: "Steven"
date: "2024-07-16"
output:
  word_document: default
  html_document: default
---
---
title: "Bike ride share - case study"
author: "Steven Roy V"
date: "2024-07-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

DATA PREPARATION

```{r}
jan=read.csv("//Users//lipikas//Downloads//cycfiles//202301-divvy-tripdata.csv")
feb=read.csv("//Users//lipikas//Downloads//cycfiles//202302-divvy-tripdata.csv")
mar=read.csv("//Users//lipikas//Downloads//cycfiles//202303-divvy-tripdata.csv")
apr=read.csv("//Users//lipikas//Downloads//cycfiles//202304-divvy-tripdata.csv")
may=read.csv("//Users//lipikas//Downloads//cycfiles//202305-divvy-tripdata.csv")
jun=read.csv("//Users//lipikas//Downloads//cycfiles//202306-divvy-tripdata.csv")
jul=read.csv("//Users//lipikas//Downloads//cycfiles//202307-divvy-tripdata.csv")
aug=read.csv("//Users//lipikas//Downloads//cycfiles//202308-divvy-tripdata.csv")
sep=read.csv("//Users//lipikas//Downloads//cycfiles//202309-divvy-tripdata.csv")
oct=read.csv("//Users//lipikas//Downloads//cycfiles//202310-divvy-tripdata.csv")
nov=read.csv("//Users//lipikas//Downloads//cycfiles//202311-divvy-tripdata.csv")
dec=read.csv("//Users//lipikas//Downloads//cycfiles//202312-divvy-tripdata.csv")
```

```{r}

library(dplyr)
# List all your dataframes
dataframes_list <- list(jan, feb, mar, apr, may, jun, jul, aug, sep, oct, nov, dec)

# Combine all dataframes into a single dataframe
combined_data <- bind_rows(dataframes_list)

# View the combined dataframe
print(combined_data)
```
```{r}
# Export the dataframe to a CSV file
write.csv(combined_data, '/Users/lipikas/Downloads/combined_data.csv', row.names = FALSE)

```

```{r}
write.csv(combined_data, '/Users/lipikas/Downloads/combined_data1.csv', row.names = FALSE)
```


```{r}
#importing data
combined_data=read.csv('//Users//lipikas//Downloads//combined_data1.csv')
combined_data
```




DATA CLEANING

```{r}
library(dplyr)
#delete columns which are not needed.
combined_data<- combined_data %>%  
  select(-c(start_station_id, end_station_id,start_lat,start_lng,end_lat,end_lng)) 
print(head(combined_data))

```
```{r}
# Count duplicates in ride_id
duplicate_count = sum(duplicated(combined_data$ride_id))
duplicate_count 
```
There are no duplicate ride id.

```{r}
#delete na values
# Replace empty strings with NA (if any)
combined_data[combined_data == ""] = NA

# Calculate NA count for each column
na_count_table = sapply(combined_data, function(col) sum(is.na(col)))

# Convert to data frame for a table view
na_count_df = data.frame(Column = names(na_count_table), NA_Count = as.integer(na_count_table))

# Print the NA count table
print(na_count_df)

na_count = sum(is.na(combined_data))
na_count # total null values

#Remove null values
combined_data = na.omit(combined_data)
print(nrow(combined_data))

```

Total number of rows = 5,719,877
A total of 1804918 null values were found and were removed.
Total number of rows after data cleaning = 4332069















DATA TRANSFORMATION

```{r}
# Convert started_at and ended_at columns to DateTime format
combined_data$started_at <- as.POSIXct(combined_data$started_at, format = "%Y-%m-%d %H:%M:%S")
combined_data$ended_at <- as.POSIXct(combined_data$ended_at , format = "%Y-%m-%d %H:%M:%S")

# Print the first few rows of the dataframe to verify
print(head(combined_data))
```
```{r}
#calculate ride length

#create new columns to show the ride length in mins and hours
combined_data$ride_length_mins <- as.numeric(difftime(combined_data$ended_at, combined_data$started_at, units = "mins"))
combined_data$ride_length_hours <- as.numeric(difftime(combined_data$ended_at, combined_data$started_at, units = "hours"))

# Round off minutes to the nearest whole number
combined_data$ride_length_mins_rounded <- round(combined_data$ride_length_mins)
# Round off hours to two decimal places
combined_data$ride_length_hours_rounded <- round(combined_data$ride_length_hours, 2)

# Print the resulting dataframe
print(combined_data)
```


```{r}
# Define intervals
intervals <- c(1000, 900, 800, 700, 600, 500, 400, 300, 200, 100, 50, 10)

# Initialize vector to store counts
counts <- numeric(length(intervals) - 1)

# Count values for each interval
for (i in seq_along(intervals[-1])) {
  counts[i] <- sum(combined_data$ride_length_mins_rounded > intervals[i + 1] & 
                    combined_data$ride_length_mins_rounded <= intervals[i], na.rm = TRUE)
}

# Print the counts
for (i in seq_along(intervals[-1])) {
  cat("Number of values between", intervals[i + 1], "and", intervals[i], ":", counts[i], "\n")
}

# Print the count for values between 0 and 10
cat("Number of values between 0 and 10:", count_between_0_and_10, "\n")
# Optionally, you can remove values within each interval if needed
# For example, to remove values between 1000 and 900:
# combined_data <- combined_data[combined_data$ride_length_mins_rounded <= 1000 & combined_data$ride_length_mins_rounded > 900, ]
# Count values less than to or equal 0
count_less_than_0 <- sum(combined_data$ride_length_mins_rounded < 0, na.rm = TRUE)
count_equal_to_0 <- sum(combined_data$ride_length_mins_rounded == 0, na.rm = TRUE)

# Count values between 0 and 10
count_between_0_and_10 <- sum(combined_data$ride_length_mins_rounded >= 0 & 
                              combined_data$ride_length_mins_rounded <= 10, na.rm = TRUE)

# Count values greater than 1000
count_greater_than_1000 <- sum(combined_data$ride_length_mins_rounded > 1000, na.rm = TRUE)

# Print the count for values less than 0
cat("Number of values less than 0:", count_less_than_0, "\n")
# Print the count for values less than 0
cat("Number of values equal to 0:", count_equal_to_0, "\n")


# Print the count for values greater than 1000
cat("Number of values greater than 1000:", count_greater_than_1000, "\n")

# Create a boxplot to visualize the distribution
library(ggplot2)
ggplot(combined_data, aes(y = ride_length_mins_rounded)) +
  geom_boxplot() +
  labs(title = "Boxplot of Ride Length (Minutes) by Interval Ranges")

```
The above mentioned values shows the ride length in mins in those specified ranges and we can see that the majority of the values 96% of the data points lie in the range 0-50 mins .
Hence i am considering the other data points as outliers and proceed with the analysis.





```{r}
nrow(combined_data)
```


DATA ANALYSIS

```{r}
#Add additional columns for analysis

library(lubridate)
library(hms)

#create column for start date of the ride
combined_data$start_date <- as.Date(combined_data$started_at) #default format is yyyy-mm-dd, use start date

#create column for day of the week start date of the ride 
combined_data$day_of_week <- wday(combined_data$started_at) 
#Sunday - 1 ... Saturday - 7

#create column for day of the week start date of the ride
combined_data$day_of_weekk <- format(as.Date(combined_data$start_date), "%A") 

#create column for month of start date of the ride
combined_data$Month <- format(as.Date(combined_data$start_date), "%m")#create column for month

#cyclistic_date$day <- format(as.Date(cyclistic_date$date), "%d") #create column for day
#cyclistic_date$year <- format(as.Date(cyclistic_date$date), "%Y") #create column for year

#cyclistic_date$time <- format(as.Date(cyclistic_date$date), "%H:%M:%S") #format time as HH:MM:SS

#create new column to find out which hour of the day was the ride started
combined_data$time <- as_hms((combined_data$started_at)) 
combined_data$hour<- hour(combined_data$time)

#group the day time to morning afternoon and evening and night

# Define time periods
combined_data$time_of_day <- with(combined_data, 
                                   case_when(
                                     hour >= 0 & hour < 6  ~ "Night",
                                     hour >= 6 & hour < 12 ~ "Morning",
                                     hour >= 12 & hour < 18 ~ "Afternoon",
                                     hour >= 18 & hour < 24 ~ "Evening"
                                   ))


#group seasons of the year as srping summer fall winter
# Create a column for different seasons
combined_data$season <- case_when(
  combined_data$Month %in% c("03", "04", "05") ~ "Spring",
  combined_data$Month %in% c("06", "07", "08") ~ "Summer",
  combined_data$Month %in% c("09", "10", "11") ~ "Fall",
  combined_data$Month %in% c("12", "01", "02") ~ "Winter"
)
print(head(combined_data))
nrow(combined_data)
```

```{r}
tabl_data = combined_data[,c('rideable_type','member_casual','ride_length_mins_rounded','ride_length_hours_rounded','day_of_week','day_of_weekk','hour','time_of_day','Month','season','start_station_name','end_station_name')]

print(head(tabl_data))
NROW((tabl_data))
```


```{r}
#write.csv(tabl_data, '/Users/lipikas/Downloads/tabl_data.csv', row.names = FALSE)

```




```{r}
# Remove extreme values  ride lengths less than or equal to 0 and greater than 50 minutes
combined_data <- combined_data[combined_data$ride_length_mins_rounded > 0 & combined_data$ride_length_mins_rounded <= 50, ]

# Print the cleaned dataframe to verify
print(head(combined_data))
NROW((combined_data))
```





DATA ANALYSIS

```{r}
#create a new dataframe for analysis with variables necessary

analysis_data = combined_data[,c('rideable_type','member_casual','ride_length_mins_rounded','ride_length_hours_rounded','day_of_week','day_of_weekk','hour','time_of_day','Month','season')]

print(head(analysis_data))
NROW((analysis_data))
```


Basic eda plots


```{r}
library(ggplot2)


# Histogram
ggplot(analysis_data, aes(x = ride_length_mins_rounded)) +
  geom_histogram(binwidth = 5, fill = "blue", color = "black") +
  labs(title = "Histogram of Ride Length (Minutes)", x = "Ride Length (Minutes)", y = "Frequency")

# Create bins for ride lengths
analysis_data$ride_length_bin <- cut(analysis_data$ride_length_mins_rounded, breaks = seq(0, max(analysis_data$ride_length_mins_rounded, na.rm = TRUE), by = 10))

# Bar plot
ggplot(analysis_data, aes(x = ride_length_bin)) +
  geom_bar(fill = "blue", color = "black") +
  labs(title = "Bar Plot of Ride Length (Minutes)", x = "Ride Length (Minutes)", y = "Count") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Density plot
ggplot(analysis_data, aes(x = ride_length_mins_rounded)) +
  geom_density(fill = "blue", alpha = 0.5) +
  labs(title = "Density Plot of Ride Length (Minutes)", x = "Ride Length (Minutes)", y = "Density")

```

```{r}
#write.csv(analysis_data, '/Users/lipikas/Downloads/analysis_data_modified.csv', row.names = FALSE)
```


Average, Min, Max, and Frequency Count of Ride Length by Month and Grouped by Casual or Member

```{r}
# Load necessary libraries
library(dplyr)
library(ggplot2)

# Sample dataframe structure
# analysis_data <- data.frame(ride_length_mins_rounded, member_casual, Month, season, day_of_week, time_of_day)

# Calculate the summary statistics
monthly_summary <- analysis_data %>%
  group_by(Month, member_casual) %>%
  summarize(
    avg_ride_length = mean(ride_length_mins_rounded, na.rm = TRUE),
    min_ride_length = min(ride_length_mins_rounded, na.rm = TRUE),
    max_ride_length = max(ride_length_mins_rounded, na.rm = TRUE),
    freq_count = n()
  )

# View the summary table
print(monthly_summary)

# Plotting the summary statistics
ggplot(monthly_summary, aes(x = Month, y = avg_ride_length, fill = member_casual)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Average Ride Length by Month and Membership Type", x = "Month", y = "Average Ride Length (mins)")

```
Average, Min, Max, and Frequency Count of Ride Length by Season and Grouped by Casual or Member


```{r}
# Calculate the summary statistics
seasonal_summary <- analysis_data %>%
  group_by(season, member_casual) %>%
  summarize(
    avg_ride_length = mean(ride_length_mins_rounded, na.rm = TRUE),
    min_ride_length = min(ride_length_mins_rounded, na.rm = TRUE),
    max_ride_length = max(ride_length_mins_rounded, na.rm = TRUE),
    freq_count = n()
  )

# View the summary table
print(seasonal_summary)

# Plotting the summary statistics
ggplot(seasonal_summary, aes(x = season, y = avg_ride_length, fill = member_casual)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Average Ride Length by Season and Membership Type", x = "Season", y = "Average Ride Length (mins)")

```
Average, Min, Max, and Frequency Count of Ride Length by Days of the Week and Grouped by Casual or Member



```{r}
# Calculate the summary statistics
day_of_week_summary <- analysis_data %>%
  group_by(day_of_week, member_casual) %>%
  summarize(
    avg_ride_length = mean(ride_length_mins_rounded, na.rm = TRUE),
    min_ride_length = min(ride_length_mins_rounded, na.rm = TRUE),
    max_ride_length = max(ride_length_mins_rounded, na.rm = TRUE),
    freq_count = n()
  )

# View the summary table
print(day_of_week_summary)

# Plotting the summary statistics
ggplot(day_of_week_summary, aes(x = day_of_week, y = avg_ride_length, fill = member_casual)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Average Ride Length by Day of Week and Membership Type", x = "Day of Week", y = "Average Ride Length (mins)")

```



Average, Min, Max, and Frequency Count of Ride Length by Time of the Day and Grouped by Casual or Member
```{r}
# Calculate the summary statistics
time_of_day_summary <- analysis_data %>%
  group_by(time_of_day, member_casual) %>%
  summarize(
    avg_ride_length = mean(ride_length_mins_rounded, na.rm = TRUE),
    min_ride_length = min(ride_length_mins_rounded, na.rm = TRUE),
    max_ride_length = max(ride_length_mins_rounded, na.rm = TRUE),
    freq_count = n()
  )

# View the summary table
print(time_of_day_summary)

# Plotting the summary statistics
ggplot(time_of_day_summary, aes(x = time_of_day, y = avg_ride_length, fill = member_casual)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Average Ride Length by Time of Day and Membership Type", x = "Time of Day", y = "Average Ride Length (mins)")

```
```{r}
library(dplyr)
library(ggplot2)

# Assuming 'analysis_data' is your dataframe

# Total count of data
total_count <- nrow(analysis_data)
print(paste("Total count of data:", total_count))

# Count of data for each month
monthly_count <- analysis_data %>%
  group_by(Month) %>%
  summarise(count = n())

print("Count of data for each month:")
print(monthly_count)

# 1. Ride Length Descriptive Statistics by Customer Type
ride_length_stats <- analysis_data %>%
  group_by(member_casual) %>%
  summarise(
    min_ride_length = min(ride_length_mins_rounded, na.rm = TRUE),
    max_ride_length = max(ride_length_mins_rounded, na.rm = TRUE),
    freq_count = n()
  )

print("Ride Length Descriptive Statistics by Customer Type:")
print(ride_length_stats)

# Plotting the frequency count of rides by customer type
ggplot(ride_length_stats, aes(x = member_casual, y = freq_count, fill = member_casual)) +
  geom_bar(stat = "identity") +
  labs(title = "Frequency Count of Rides by Customer Type", x = "Customer Type", y = "Frequency Count") +
  theme_minimal()

# 2. Monthly Analysis by Customer Type
monthly_stats <- analysis_data %>%
  group_by(Month, member_casual) %>%
  summarise(
    min_ride_length = min(ride_length_mins_rounded, na.rm = TRUE),
    max_ride_length = max(ride_length_mins_rounded, na.rm = TRUE),
    freq_count = n()
  )

print("Monthly Analysis by Customer Type:")
print(monthly_stats)

# Plotting the frequency count of rides by month and customer type
ggplot(monthly_stats, aes(x = Month, y = freq_count, fill = member_casual)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Frequency Count of Rides by Month and Customer Type", x = "Month", y = "Frequency Count") +
  theme_minimal()

# 3. Seasonal Analysis by Customer Type
seasonal_stats <- analysis_data %>%
  group_by(season, member_casual) %>%
  summarise(
    min_ride_length = min(ride_length_mins_rounded, na.rm = TRUE),
    max_ride_length = max(ride_length_mins_rounded, na.rm = TRUE),
    freq_count = n()
  )

print("Seasonal Analysis by Customer Type:")
print(seasonal_stats)

# Plotting the frequency count of rides by season and customer type
ggplot(seasonal_stats, aes(x = season, y = freq_count, fill = member_casual)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Frequency Count of Rides by Season and Customer Type", x = "Season", y = "Frequency Count") +
  theme_minimal()

# 4. Day of the Week Analysis by Customer Type
day_stats <- analysis_data %>%
  group_by(day_of_week, member_casual) %>%
  summarise(
    min_ride_length = min(ride_length_mins_rounded, na.rm = TRUE),
    max_ride_length = max(ride_length_mins_rounded, na.rm = TRUE),
    freq_count = n()
  )

print("Day of the Week Analysis by Customer Type:")
print(day_stats)

# Plotting the frequency count of rides by day of the week and customer type
ggplot(day_stats, aes(x = day_of_week, y = freq_count, fill = member_casual)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Frequency Count of Rides by Day of the Week and Customer Type", x = "Day of the Week", y = "Frequency Count") +
  theme_minimal()

# 5. Time of Day Analysis by Customer Type
time_stats <- analysis_data %>%
  group_by(time_of_day, member_casual) %>%
  summarise(
    min_ride_length = min(ride_length_mins_rounded, na.rm = TRUE),
    max_ride_length = max(ride_length_mins_rounded, na.rm = TRUE),
    freq_count = n()
  )

print("Time of Day Analysis by Customer Type:")
print(time_stats)

# Plotting the frequency count of rides by time of day and customer type
ggplot(time_stats, aes(x = time_of_day, y = freq_count, fill = member_casual)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Frequency Count of Rides by Time of Day and Customer Type", x = "Time of Day", y = "Frequency Count") +
  theme_minimal()

```

```{r}
library(dplyr)
library(ggplot2)
library(scales)

# Assuming 'analysis_data' is your dataframe

# Total count of data
total_count <- nrow(analysis_data)
print(paste("Total count of data:", total_count))

# Count of data for each month
monthly_count <- analysis_data %>%
  group_by(Month) %>%
  summarise(count = n())

print("Count of data for each month:")
print(monthly_count)

# 1. Ride Length Descriptive Statistics by Customer Type
ride_length_stats <- analysis_data %>%
  group_by(member_casual) %>%
  summarise(
    min_ride_length = min(ride_length_mins_rounded, na.rm = TRUE),
    max_ride_length = max(ride_length_mins_rounded, na.rm = TRUE),
    freq_count = n()
  )

print("Ride Length Descriptive Statistics by Customer Type:")
print(ride_length_stats)

# Plotting the frequency count of rides by customer type
ggplot(ride_length_stats, aes(x = member_casual, y = freq_count, fill = member_casual)) +
  geom_bar(stat = "identity") +
  scale_y_continuous(labels = comma) +
  labs(title = "Frequency Count of Rides by Customer Type", x = "Customer Type", y = "Frequency Count") +
  theme_minimal()

# 2. Monthly Analysis by Customer Type
monthly_stats <- analysis_data %>%
  group_by(Month, member_casual) %>%
  summarise(
    min_ride_length = min(ride_length_mins_rounded, na.rm = TRUE),
    max_ride_length = max(ride_length_mins_rounded, na.rm = TRUE),
    freq_count = n()
  )

print("Monthly Analysis by Customer Type:")
print(monthly_stats)

# Plotting the frequency count of rides by month and customer type
ggplot(monthly_stats, aes(x = Month, y = freq_count, fill = member_casual)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_y_continuous(labels = comma) +
  labs(title = "Frequency Count of Rides by Month and Customer Type", x = "Month", y = "Frequency Count") +
  theme_minimal()

# 3. Seasonal Analysis by Customer Type
seasonal_stats <- analysis_data %>%
  group_by(season, member_casual) %>%
  summarise(
    min_ride_length = min(ride_length_mins_rounded, na.rm = TRUE),
    max_ride_length = max(ride_length_mins_rounded, na.rm = TRUE),
    freq_count = n()
  )

print("Seasonal Analysis by Customer Type:")
print(seasonal_stats)

# Plotting the frequency count of rides by season and customer type
ggplot(seasonal_stats, aes(x = season, y = freq_count, fill = member_casual)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_y_continuous(labels = comma) +
  labs(title = "Frequency Count of Rides by Season and Customer Type", x = "Season", y = "Frequency Count") +
  theme_minimal()

# 4. Day of the Week Analysis by Customer Type
day_stats <- analysis_data %>%
  group_by(day_of_week, member_casual) %>%
  summarise(
    min_ride_length = min(ride_length_mins_rounded, na.rm = TRUE),
    max_ride_length = max(ride_length_mins_rounded, na.rm = TRUE),
    freq_count = n()
  )

print("Day of the Week Analysis by Customer Type:")
print(day_stats)

# Plotting the frequency count of rides by day of the week and customer type
ggplot(day_stats, aes(x = day_of_week, y = freq_count, fill = member_casual)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_y_continuous(labels = comma) +
  labs(title = "Frequency Count of Rides by Day of the Week and Customer Type", x = "Day of the Week", y = "Frequency Count") +
  theme_minimal()

# 5. Time of Day Analysis by Customer Type
time_stats <- analysis_data %>%
  group_by(time_of_day, member_casual) %>%
  summarise(
    min_ride_length = min(ride_length_mins_rounded, na.rm = TRUE),
    max_ride_length = max(ride_length_mins_rounded, na.rm = TRUE),
    freq_count = n()
  )

print("Time of Day Analysis by Customer Type:")
print(time_stats)

# Plotting the frequency count of rides by time of day and customer type
ggplot(time_stats, aes(x = time_of_day, y = freq_count, fill = member_casual)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_y_continuous(labels = comma) +
  labs(title = "Frequency Count of Rides by Time of Day and Customer Type", x = "Time of Day", y = "Frequency Count") +
  theme_minimal()

```

